---
title: "Using promptbook with ellmer directly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using promptbook with ellmer directly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `pb_annotate()` convenience function handles the full annotation
pipeline. But sometimes you need more control --- custom prompts per row,
different extraction strategies, or integration with your own workflow.
This vignette shows how to use `pb_type()` to get an ellmer type object
and work with it directly.

## From YAML to ellmer type

`pb_type()` converts a promptbook (or a subset of its variables) into
an `ellmer::type_object()`. This is the bridge between the YAML schema
and ellmer's structured data extraction.

```{r}
library(promptbook)

path <- system.file("examples", "media_framing.yaml", package = "promptbook")
pb <- read_promptbook(path)

# Convert all variables to an ellmer type
type_all <- pb_type(pb)
type_all
```

### Subsetting by group

Extract only variables from a specific group:

```{r}
type_basic <- pb_type(pb, group = "basic")
type_basic
```

### Subsetting by name

Pick specific variables:

```{r}
type_sub <- pb_type(pb, variables = c("topic", "sentiment"))
type_sub
```

## Using with ellmer directly

Once you have a type object, use it with any ellmer chat method:

```{r, eval = FALSE}
library(ellmer)

chat <- chat_anthropic(
  model = "claude-sonnet",
  system_prompt = pb$prompt$system
)

# Single text
result <- chat$chat_structured(
  "The government announced a $2 billion climate fund...",
  type = type_basic
)
result

# Multiple texts in parallel
texts <- list(
  "Article about economic impacts...",
  "Article about health policy...",
  "Article about renewable energy..."
)

results <- parallel_chat_structured(chat, texts, type = type_basic)
results
```

## Type mapping reference

Each YAML variable type maps to an ellmer type function. The variable's
`description` field is enriched with type-specific information (categories,
scale endpoints, etc.) and passed as the type description:

```{r, echo = FALSE}
mapping <- data.frame(
  YAML = c("categorical", "categorical (multiple)", "numeric",
           "numeric (integer: false)", "text", "text (multiple)",
           "boolean", "object", "object (multiple)"),
  ellmer = c("type_enum()", "type_array(type_enum())", "type_integer()",
             "type_number()", "type_string()", "type_array(type_string())",
             "type_boolean()", "type_object()", "type_array(type_object())")
)
knitr::kable(mapping, col.names = c("YAML type", "ellmer function"))
```

## Example: custom per-row prompts

One reason to use `pb_type()` directly is when each row needs a different
prompt --- for example, if context varies across observations:

```{r, eval = FALSE}
library(ellmer)

type <- pb_type(pb, group = "basic")

chat <- chat_anthropic(
  model = "claude-haiku",
  system_prompt = pb$prompt$system
)

results <- lapply(seq_len(nrow(articles)), function(i) {
  # Build a custom prompt for each row
  prompt <- sprintf(
    "Country context: %s\n\nArticle:\n%s",
    articles$country[i],
    articles$text[i]
  )
  cloned <- chat$clone()
  cloned$chat_structured(prompt, type = type)
})

do.call(rbind, results)
```

## Example: two-stage annotation

Another use case is running a fast model first, then a stronger model
only for uncertain cases:

```{r, eval = FALSE}
library(ellmer)

# Stage 1: fast model on everything
type_basic <- pb_type(pb, group = "basic")

chat_fast <- chat_anthropic(model = "claude-haiku",
                            system_prompt = pb$prompt$system)
stage1 <- parallel_chat_structured(chat_fast, texts, type = type_basic)

# Stage 2: strong model only where sentiment is neutral (uncertain)
uncertain <- which(stage1$sentiment == 3)

if (length(uncertain) > 0) {
  chat_strong <- chat_anthropic(model = "claude-sonnet",
                                system_prompt = pb$prompt$system)
  stage2 <- parallel_chat_structured(
    chat_strong, texts[uncertain], type = type_basic
  )
  stage1[uncertain, ] <- stage2
}
```

## The pb_var_to_type() generic

For even finer control, `pb_var_to_type()` converts a single variable
to its ellmer type. This is the S3 generic that `pb_type()` uses
internally:

```{r}
# Get a single variable
var <- pb$variables[[1]]
var

# Convert to ellmer type
ellmer_type <- pb_var_to_type(var)
ellmer_type
```

This is useful if you need to compose type objects manually or inspect
how a specific variable maps to ellmer.
