---
title: "Getting started with promptbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with promptbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The promptbook package helps researchers annotate text data with LLMs in a
documented, reproducible way. You define your coding instrument once as a
YAML file --- the **promptbook** --- and the package handles turning it into
structured LLM prompts, a human-readable codebook, and properly typed R output.

## The YAML file

A promptbook YAML file is the single source of truth for your coding
instrument. It contains everything needed to generate both a human codebook
and structured LLM prompts.

Here's a minimal example:

```yaml
schema_version: 1
title: "Sentiment Codebook"
author: "Jane Researcher"

prompt:
  system: |
    You are an expert content analyst. You will be given a text
    and must code it according to the variables below.

variables:
  - name: sentiment
    label: "Overall sentiment"
    description: >
      The overall evaluative tone of the text.
      Use the full scale; reserve 3 for genuinely neutral texts.
    type: numeric
    min: 1
    max: 5
    labels:
      1: "Very negative"
      2: "Somewhat negative"
      3: "Neutral"
      4: "Somewhat positive"
      5: "Very positive"
```

### Top-level fields

| Field | Required | Description |
|---|---|---|
| `schema_version` | yes | Integer; currently `1`. For forward compatibility. |
| `title` | yes | Human-readable title for the coding instrument. |
| `version` | no | Semantic version for the codebook (e.g., `"1.2.0"`). |
| `description` | no | Longer description; appears in the rendered codebook. |
| `author` | no | Author name(s). |

### The `prompt` section

| Field | Required | Description |
|---|---|---|
| `system` | yes | System prompt sent to the LLM. |
| `user` | no | User message template with `{text}` placeholder. Defaults to `"{text}"` if omitted. |

The system prompt is your main lever for controlling LLM behavior. Variable
descriptions are embedded automatically in the structured output schema,
so you don't need to repeat them in the prompt.

If your data has multiple text columns (e.g., title and abstract), use
named placeholders:

```yaml
prompt:
  user: |
    Title: {title}
    Abstract: {abstract}
```

### Variable types

Each variable in the `variables` list defines one code to assign. Every
variable needs `name`, `type`, and `description`. The five types are:

**`categorical`** --- pick one (or many) from a set of categories:

```yaml
- name: topic
  label: "Primary topic"
  description: "The dominant topic of the article."
  type: categorical
  categories:
    - value: "economy"
      label: "Economic impacts"
      description: "Focus on costs, jobs, trade effects"
    - value: "health"
      label: "Public health"
    - value: "environment"
      label: "Environmental impacts"
```

Add `multiple: true` for "select all that apply" questions.

**`numeric`** --- an integer or continuous scale:

```yaml
- name: sentiment
  description: "Overall tone. Use the full scale."
  type: numeric
  min: 1
  max: 5
  labels:
    1: "Very negative"
    5: "Very positive"
```

Defaults to integer; set `integer: false` for continuous values.

**`text`** --- free-form text response:

```yaml
- name: key_quote
  description: "The most representative direct quote."
  type: text
```

**`boolean`** --- yes/no:

```yaml
- name: has_data
  description: "Whether the article cites quantitative data."
  type: boolean
```

**`object`** --- structured records (e.g., extracting actors):

```yaml
- name: actors
  description: "All political actors quoted in the article."
  type: object
  multiple: true
  properties:
    - name: actor_name
      type: text
      description: "Full name"
    - name: stance
      type: categorical
      description: "Stance on the policy"
      categories:
        - value: "support"
        - value: "oppose"
        - value: "neutral"
```

### Common variable fields

| Field | Default | Description |
|---|---|---|
| `name` | (required) | Column name in output. Must be a valid R name. |
| `label` | same as `name` | Human-readable label for the codebook. |
| `description` | (required) | Sent to the LLM *and* printed in the codebook. |
| `type` | (required) | One of: `categorical`, `numeric`, `text`, `boolean`, `object`. |
| `required` | `false` | If `true`, forces the LLM to always produce a value. Keep `false` to avoid hallucination when data doesn't fit. |
| `multiple` | `false` | If `true`, returns an array. Valid for `categorical`, `text`, `object`. |
| `group` | (none) | Group name for batching variables into separate LLM calls. |
| `model` | (none) | Model override for this variable (or set at the group level). |

### Groups

When you have many variables or want to use different models for different
codes, define groups:

```yaml
variables:
  - name: topic
    # ...
    group: basic

  - name: frame
    # ...
    group: framing

groups:
  basic:
    label: "Basic codes"
    model: fast
  framing:
    label: "Framing analysis"
    model: strong
```

Variables in the same group are extracted in one LLM call. Each group can
use a different model --- fast/cheap models for simple codes, stronger
models for complex ones.

## Reading a promptbook

```{r}
library(promptbook)

path <- system.file("examples", "media_framing.yaml", package = "promptbook")
pb <- read_promptbook(path)
pb
```

All validation happens in `read_promptbook()`. If it returns without error,
the YAML is valid. You can inspect the structure:

```{r}
pb$title
pb$variables[[1]]
```

## Annotating data

`pb_annotate()` is the main workhorse. It takes your data, the promptbook,
and one or more [ellmer](https://ellmer.tidyverse.org) chat objects:

```{r, eval = FALSE}
library(ellmer)

# Create chat objects for each model used in the promptbook
chat <- list(
  fast   = chat_anthropic(model = "claude-haiku"),
  strong = chat_anthropic(model = "claude-sonnet")
)

# Your data --- must have a column matching the {text} placeholder
articles <- data.frame(
  id   = 1:3,
  text = c("Article about economic policy...",
           "Article about health impacts...",
           "Article about new technology...")
)

# Annotate!
result <- pb_annotate(articles, pb, chat)
result
```

The result is a tibble with your original columns plus new columns for each
promptbook variable, properly typed (factors for categorical, integers for
numeric, etc.).

### Dispatch methods

`pb_annotate()` supports three methods:

- `"parallel"` (default) --- sends all prompts concurrently. Fast.
- `"batch"` --- uses the batch API. 50% cheaper, but slow (up to 24h).
- `"sequential"` --- one prompt at a time. Useful for debugging.

```{r, eval = FALSE}
result <- pb_annotate(articles, pb, chat, method = "sequential")
```

### Multiple text columns

If your data has multiple text columns, use `{placeholder}` syntax in the
YAML user template and select the columns in `pb_annotate()`:

```{r, eval = FALSE}
# YAML has: user: "Title: {title}\nAbstract: {abstract}"
result <- pb_annotate(articles, pb, chat, text = c(title, abstract))
```

## Rendering a codebook

Generate a human-readable codebook from the same YAML file:

```{r, eval = FALSE}
pb_render(pb, "codebook.html")
```

This produces a self-contained HTML document with all variable definitions,
categories, and the system prompt --- suitable for paper appendices or
sharing with human coders. PDF output via Typst is also supported:

```{r, eval = FALSE}
pb_render(pb, "codebook.pdf", format = "typst")
```

## Exporting to SPSS/Stata

For researchers who work with SPSS or Stata, convert annotation results
to haven-labelled format:

```{r, eval = FALSE}
labelled_result <- pb_as_labelled(result, pb)
haven::write_sav(labelled_result, "coded_articles.sav")
```

Categorical variables become integer-coded with value labels, booleans
become 0/1, and variable labels are attached from the YAML.

## A complete example

Here's the full media framing example that ships with the package:

```{r, echo = FALSE, comment = ""}
cat(readLines(path), sep = "\n")
```

This promptbook defines 8 variables in 2 groups, uses different models
for simple vs. complex codes, and covers all five variable types including
an `object` type for extracting structured actor records.
